Two Problems for Sophistication

### 2015-05-12 Facticity/Facticity/paper.tex
Ugh...

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\bi}{\bar\imath}
\newcommand{\dom}[1]{\mathop{\tn{dom}(#1)}}
\newcommand{\range}[1]{\mathop{\tn{range}(#1)}}

@@

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of the smallest model allowing for optimal compression of the data. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe two fundamental problems with existing proposals, showing many of them to be unsound. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

\section{Introduction}

Kolmogorov complexity gives us a sound definition of the amount of information contained in a binary string. It does not, however, capture what most people would consider complexity. For example, a sequence of a million coin flips will almost certainly have maximal Kolmogorov complexity, even though there is nothing complex about flipping a coin repeatedly. Many scholars have defined additional complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information. While this concept has been given many names, we use \emph{sophistication} as an umbrella term. In this paper, we investigate two serious problems with sophistication measures. We conclude with three arguments that suggest the problems are fundamental, explaining our belief that sophistication cannot be defined in a satisfactory manner.

The Kolmogorov complexity $C(x)$ of a binary string $x$, is (roughly) the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but, as shown by a central result known as the invariance theorem \cite[Section~2.1]{li1993introduction}, the language affects the value only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings, a property \emph{of the data.} A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except that it's also bounded). Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, which is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

For any string $x$, there may be many different two-part representations. The total codelength can never be less than the Kolmogorov complexity, but it can come close. The balance between the two parts may vary; Figure~\ref{fig:diagram} provides an illustration. The various proposals then define sophistication in terms of the length of the structure part of this two-part code for one of the representations that (almost) achieves the Kolmogorov complexity. However, for most of these definitions, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we show that this method only assigns substantial sophistication to strings that require an enormous amount of processing to construct.

\begin{figure*}[tb]
  \centering
  \begin{minipage}{0.45\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.45\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$ by the two components of their codelength. The Kolmogorov complexity $C^\M(x)$, appearing as a black diagonal, provides a lower bound on the total codelength. We consider only representations that are close to this optimum, with the threshold represented by a dashed line. The size of the smallest model below the threshold is the sophistication of the data. (right) The same image, after a constant perturbation in the model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}

@@

The second issue is that of striking the right balance between underfitting and overfitting, which we consider in Section~\ref{section:balance}. Overfitting is a common problem in the statistical literature, that refers to the tendency of model selection to choose a complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. In the case of sophistication, overfitting occurs if the model that determines the sophistication contains much or even all of the noise. This is a real danger, because the models are encoded so efficiently. In statistics, overfitting is often addressed by penalising complex models. In sophistication, however, such penalties tend to break the balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting occurs when the selected model is simple, but fails to represent some of the structure present in the data. This is also a problem for sophistication because the models under consideration are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} are \emph{simple} in the language of sophistication, since they can be described with a relatively small number of bits, yet are able to represent the data within a codelength a constant away from the Kolmogorov complexity. Such a two-part representation essentially encodes all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions. While this excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, in the discussion in Section~\ref{section:conclusion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as poor representations, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable representations. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that the data allows for multiple equivalent interpretations of which information is structured, and which is random, and that there is no such thing as sophistication.

\section{Preliminaries}
One of the main issues in dealing with sophistication in a general sense, is that one must unify many different approaches, all founded on slightly different principles. We have devised a general framework which, while it may not represent the idiosyncracies of all methods, allows us to rephrase them in a uniform way that illustrates the properties we discuss. We hope the reader will trust that these properties remain also in the original statements of the measures.

Let $\B = \{0,1\}^*$. We deal with partial computable functions $f: \B \times \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\dom{f} = \{y : f(y; \cdot) \neq \infty\}$ is a prefix free set, i.e. no string in $\dom{f}$ is a prefix of another. A function $f$ is \emph{total} if $\dom{f} = \B$. In most cases, we do not use the second argument, and we adopt the convention that $f(x) = f(x, \epsilon)$.

@@

A \emph{model class} is a set of model indices. We distinguish four important model classes:
\begin{itemize} 
  \item The indices of the partial computable functions $\C=\N$.
  \item The total functions $\T=\{i:\tn{$\phi_i$ is total}\}$. Note that $\T$ is not computably enumerable.
  \item $\K$ is an enumerable set such that $\{\phi_i:i\in\K\}$ is the set of all partial computable prefix functions.
  \item The finite sets: $\F$ is an enumerable set such that $\{\phi_i:i\in\F\}$ contains a uniform code for every finite set. A uniform code for a finite set $S$ is a total computable surjective function $f:\{0,1\}^{\lceil\log|S|\rceil}\to S$.
\end{itemize}

@@

For technical reasons, we deviate slightly from the traditional notation of Kolmogorov complexity: we define the $\M$-Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows $C^{\M,\psi}(x\mid z)=\min\{\bar\imath y:\psi_i(y; z)=x,i\in\M\}$, with $C^\M(x) = C^\M(x\mid \epsilon)$. This means that $C^\C(x)$ corresponds to the plain Kolmogorov complexity, traditionally denoted $C(x)$, and $C^\K(x)$ corresponds to the prefix-free Kolmogorov complexity $K$. This notation also allows us to represent the smallest possible description of $x$ using a single model $\phi_i$ as $C^{\{i\}}(x)$. Note that this includes the description of $i$ as well as its shortest program for $x$.

Finally we note that the choice of numbering, combined with a choice of prefix function, serves the same purpose that is normally expressed in the choice of a Turing machine. Requiring an acceptable numbering and a computable prefix function is equivalent to choosing a universal Turing machine: this requirement suffices to make the Kolmogorov complexity invariant. As we shall see, however, this is not enough for a proper treatment of sophistication: we must choose our numbering with care. Thus, we phrase our requirement for an invariant sophistication with respect to the \emph{numbering} chosen. There should be strict guarantees on how much the value of $S(x)$ can change if the numbering is varied.

@@

The simplest approach to sophistication would simply be to `open up' the Kolmogorov complexity and to see which program on the universal Turing machine achieves the smallest description length: the program that \emph{witnesses} the Kolmogorov complexity. Since any UTM is defined in terms of a two-part coding, such programs necessarily consist of a model and an input. As is common, we will allow all witnesses within a constant slack term $c$ from the Kolmogorov complexity, and choose the one with the smallest model.

\begin{definition}[Index sophistication]
\[
\s_{\text{index}}^{\M_1,\M_2,\psi,c}(X) = \min\left\{ |i| : \i \in \M_1, C^{\{i\}, \psi}(x) \leq C^{\M_2, \psi}(x)+c   \right\}
\]
When $\M_1 = \M_2$, we will abbreviate with $\s^{\M_1,\psi,c}$
\end{definition}

Koppel and Atlan's treatment \cite{koppelSoph1988,koppel1991almost}, where the name \emph{sophistication} originates, follows this basic logic, although it contains many idiosyncracies like the use of monotonic models, and an inherent extension to infinite strings. As the subsequent history of sophistication has discarded these, we will not discuss them here.

In \cite{antunes2009sophistication,antunes2013sophistication} Koppel's principle is recast to finite strings. The definition used is similar to $s_\text{index}^{\T, \C,\psi,c}$, with one exception: the total complexity of a witness $(i,y)$ is measured as $|i|+|y|$ without the cost of delimiting the two. This difference is not relevant to the point we make in this section. The restriction to total classes is a common technique to avoid underfitting, which we will discuss in the next section.

\begin{lemma}
Let $s^\psi$ denote any index sophistication with respect to numbering $\psi$.
There are acceptable numberings $\psi$ and $\xi$ such that for all $x$:\\
\-\hspace{2.8em} $|s^\psi(x) - s^\xi(x)|\geq \tfrac{1}{2}\min\{s^\psi(x),s^\xi(x)\}$.
\end{lemma}
\begin{proof}
Let $s_i$ be the string consisting of $2^{i}$ zeroes followed by a one. Define $\phi, \xi$ such that $\psi_j(x) = \phi_i(x)$ for $j = s_{2i}$ and $\xi_j(x) = \phi_i(x)$ for $j = s_{2i+1}$, with all other functions returning $\infty$. Choose any $x$ and assume w.l.o.g. that $s^\psi(x)\le s^\xi(x)$. By construction, we have $2s_\psi(x)  \leq s_\xi(x)$ from which the lemma follows.
\end{proof}
Thus, the length of the index is a very poor indicator of model complexity. In order to obtain a robust measure, define the complexity of functions as in \cite{grunwald2004shannon,vitanyi2004meaningful} by $C^{\M,\psi}(f) = \min\{C^{\M,\psi}(i):\phi_i=f\}$.
Lemma~\ref{lemma:invariance} in the appendix shows that this definition is invariant: it is an objective property of a function.

Note that such perversely inefficient numberings are not an issue for Kolmogorov complexity. If our universal Turing machine uses an inefficient numbering, we can switch to a more efficient one at only a constant cost. This gives us an insight into what happens inside Kolmogorov complexity. We may define the program $\bi y$ using a model $i$ and an input $y$, but if there's any inefficiency in the description of $i$, the shortest program will be achieved by using $i$ to define a more efficient model encoding, and encoding both the model and its input in $y$. For the Kolmogorov complexity, such considerations are academic, since we only look at the total length of the program. For sophistication, however, they are crucial.

There are two ways to use the model complexity $C^\M(\phi_i)$ for more robust attempts to define sophistication. Confusingly, both are used in the literature. First, we can measure the complexity of the \emph{model} using $C^{\M,\psi}(\phi_i)$, which is then the first part of a two-part code describing the data. This approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information} and in most of the present paper.

Second, we can stick to using the length of the index as the measure of sophistication, but restrict the allowed numberings. This approach is taken is in \cite{adriaans2012facticity}. Adriaans  defines \emph{facticity} as $\s_\text{index}^{\C,\psi,0}$, but only allows numberings that are \emph{faithful}: for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and $|j|\le C^{\C}(\psi_j)+c$, for some constant $c$.

We prove in the appendix that---contrary to Adriaans' suggestion---there actually do exist faithful, acceptable numberings (Lemma~\ref{lemma:faithful-numberings}). However even choosing a faithful index is not quite enough. The Kolmogorov complexity uses representations of the form $\bar\imath y$, where the bar denotes some straightforward prefix encoding to delimit the model description $i$ from its input $y$. If we define a second prefix encoding $\tilde{\imath}$, with $|\bar\imath|-|\tilde\imath|$ unbounded, we can define a second representation $\bar u \tilde \imath y$, at a constant overhead $|\br{u}|$, and gain more than $|\br{u}|$ for sufficiently complex strings. As shown in Lemma~\ref{lemma:prefix-inefficiency} in the appendix, this results in a bounded sophistication (thus the facticity is bounded).

Because of these issues, rather than index sophistication we will use the following definition of sophistication in the remainder of this article. It is parameterised by a model class, a numbering, and a slack parameter $c$ that determines the maximal allowed slack between the two-part code length using a representation and the Kolmogorov complexity.
\[
\s^{\M_1,\M_2,\psi,c}(x)=\min\left\{C^\K(\phi_i):i\in\M_1,C^{\{i\},\psi}(x)\le C^{\M_2,\psi}(x)+c\right\}.
\]

@@

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. We will investigate this balance, starting with the variant $\s^{\K,\psi,c}$. For reasons which will become clear, this variant is not used anywhere in the literature, but it helps to illustrate the issues we wish to discuss.

In this setting, we know that, on the one hand, there always exist witnesses for which all information except a constant amount is stored in the input, and on the other hand, there are also candidate representations for which all information is stored in the model. This tells us that $S^\K$ is \emph{balanced}: there is no inefficiency in the representation forcing information in or out of the model.
The flip side of having a balanced sophistication is that it becomes easy to show a lack of invariance, by tweaking the numbering in such a way that models in a specific subset $\M'\subset\M$ of the model class becomes an arbitrary constant number of bits cheaper to represent than all other models, so that a model in $\M'$ always ends up determining the sophistication.

This means that for model classes with a universal model, we can choose our numbering to make $S^\K$ bounded, or equal to the complexity:

\begin{restatable}[Underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M : C^{\{u\}}(x) \leq C^{\{i\}}(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.\label{lemma:underfitting}
\end{restatable}

For this reason, many treatments restrict the model class. What is perhaps less well known is that the same property holds in the other direction: there are numberings for which the singleton models always determine the sophistication.

\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\M(x)-S^{\M,\psi}(x)\leq c$.\label{lemma:overfitting}
\end{restatable}

The proof of both these lemmas relies on a simple principle: for certain subclasses $\M' \subset \M$, there exist numberings which, when used, have the effect of penalizing the codelength $C^\K(\phi_i)$ for any model outside $\M'$ by an arbitrary constant amount. We can use this penalty to effectively `push' these models outside of the range of considered models, ensuring that, under this numbering, a model in $\M'$ always determines the sophistication.

For the lemmas above, this means setting $\M' = {u}$, with $\phi_u$ a universal model, or letting $\M'$ be the set of all singletons. The requirements for $\M'$ are strict, and slightly complex. The following lemma gives a set of sufficient conditions.

>>>>

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\Nm}{\mathscr N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\bi}{\bar\imath}
\newcommand{\dom}[1]{\mathop{\tn{dom}(#1)}}
\newcommand{\range}[1]{\mathop{\tn{range}(#1)}}

@@

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of the smallest model that allows for optimal compression of the data. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe two fundamental problems with existing proposals, showing many of them to be unsound. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

\section{Introduction}
\enlargethispage{2\baselineskip}
\footnotetext{Peter Bloem is a student author (PhD student).}

Kolmogorov complexity gives us a sound definition of the amount of information contained in a binary string. It does not, however, capture what most people would consider complexity. For example, a sequence of a million coin flips will almost certainly have maximal Kolmogorov complexity, even though there is nothing complex about flipping a coin repeatedly. Many scholars have defined additional measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful}. While this concept has been given many names, we use \emph{sophistication} as an umbrella term. In this paper, we investigate two serious problems with sophistication. We conclude with two arguments suggesting the problems are fundamental, explaining our belief that sophistication cannot be defined in a satisfactory manner.

The Kolmogorov complexity $C(x)$ of a binary string $x$ is, informally, the length of the shortest computer program to print $x$. This length depends on the choice of programming language, but, by the invariance theorem \cite[Section~2.1]{li1993introduction}, this choice affects the value only by a constant, independent of $x$. For sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure. A definition of sophistication $S(x)$ in the spirit of $C(x)$ should have similar guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except it is also bounded). 
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many proposals for such a measure, all based on a \emph{two-part code}: we encode a \emph{model} in the first part of the code, which is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined.\footnotemark

\footnotetext{Some variants deviate from the two-part coding format, see Section~\ref{section:other}.}

For any string $x$, there may be many different two-part codes. The total length can never be less than the Kolmogorov complexity, but it can come close. Figure~\ref{fig:diagram} illustrates the principle. The various proposals define sophistication as the size of the model for one of the representations that approaches the Kolmogorov complexity. However, for most definitions, we can prove  they fail one of the conditions above. For certain variants, we cannot \emph{prove} they conflict with our requirements, but we show these methods only assign substantial sophistication to strings that require an enormous amount of processing to construct.

\begin{figure*}[tb]
  \centering
  \begin{minipage}{0.40\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.40\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$ by the two components of their codelength. The Kolmogorov complexity $C^\M(x)$, appearing as a black diagonal, provides a lower bound on the total codelength. We consider only representations that are close to this optimum, with the threshold represented by a dashed line. The size of the smallest model below the threshold is the sophistication of the data. (right) The same image, after a constant perturbation in the model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}

@@

The second issue is that of striking the right balance between under- and overfitting, which we consider in Section~\ref{section:balance}. Overfitting is a common problem in statistics, that refers to the tendency to choose a complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. In the case of sophistication, overfitting occurs if the model that determines the sophistication contains much or even all of the noise. In statistics, overfitting is often addressed by penalising complex models. In sophistication, however, such penalties tend to break the balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting occurs when the selected model is simple, but fails to represent some of the structure present in the data. This is also a problem for sophistication because the models under consideration are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} are \emph{simple}, since they can be described with a relatively small number of bits, yet are able to represent the data using a code within a constant from the Kolmogorov complexity. Such a two-part representation essentially encodes all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models, for instance to total functions. While this excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, In the discussion in Section~\ref{section:conclusion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as poor representations, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable representations. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that the data allows for multiple equivalent interpretations of which information is structured, and which is random, and that there is no such thing as sophistication.

\enlargethispage{3\baselineskip} 
\section{Notation}
The following notation allows us to generalize across all definitions and variants, save the occasional exception which we will highlight individually.

Let $\B = \{0,1\}^*$. We deal with partial computable functions $f: \B \times \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\text{dom}_z(f)~=~\{y~:~f(y,~z)~\neq \infty\}$, is a prefix free set for all $z$, i.e. no string in $\text{dom}_z(f)$ is a prefix of another. A function $f$ is \emph{total} if $\forall_z \text{dom}_z(f) = \B$. In most cases, we do not use the second argument, and we adopt the convention that $f(x) = f(x, \epsilon)$.

@@

A \emph{model class} is a set of model indices. We distinguish four classes:
\begin{itemize} 
  \item The indices of the partial computable functions $\C=\N$.
  \item The total functions $\T=\{i:\tn{$\phi_i$ is total}\}$. Note that $\T$ is not computably enumerable.
  \item $\K$ is an enumerable set such that $\{\phi_i:i\in\K\}$ is the set of all partial computable prefix functions.
  \item The finite sets: $\F$ is an enumerable set such that $\{\phi_i:i\in\F\}$ contains a uniform code for every finite set.\footnote{A uniform code for a set $S$ is a surjective prefix function $f:\{0,1\}^{\lceil\log|S|\rceil}\to S$.}
\end{itemize}
A model class is thus defined in reference to a specific numbering, indicated by the context.

@@

For technical reasons, we deviate slightly from the traditional notation of Kolmogorov complexity: we define the $\M$-Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows $C^{\M,\psi}(x\mid z)=\min\{\bar\imath y:\psi_i(y; z)=x,i\in\M\}$, with $C^{\M,\psi}(x) = C^{\M,\psi}(x\mid \epsilon)$. We may drop the numbering parameter when the distinction isn't relevant. Under this notation $C^\C(x)$ corresponds to the plain Kolmogorov complexity $C(x)$ and $C^\K(x)$ corresponds to the prefix-free version $K(x)$. We can also represent the smallest two-part description of $x$ using a given model $\phi_i$ as $C^{\{i\}}(x)$. The complexity of a finite set is defined an the complexity of an explicit enumeration.

We use a choice of numbering for the purpose that is normally served by the universal Turing machine. The choice is a matter of taste: a UTM determines a numbering, and vice versa. The reason that we prefer to work with numberings is that it highlights an important issue: while Kolmogorov complexity is invariant to the choice of numbering this property does not immediately carry over to sophistication: for some treatments, the result is highly dependent on the chosen numbering (as shown in the next section).

@@

The simplest approach to sophistication would be to `open up' the Kolmogorov complexity and to see which program on the universal Turing machine achieves the smallest description length: the program that \emph{witnesses} the Kolmogorov complexity. Since any UTM uses a two-part coding, its programs consist of a model and an input:

\begin{definition}[Index sophistication]
Let $\psi$ be an acceptable numbering. Let $\M$ be the model class from which candidates are chosen, and let $\Nm$ be the model class that determines the minimum achievable complexity. Let $c$ be a fixed constant. The \emph{index sophistication} is:
\[
\s_{\text{index}}^{\M,\Nm,\psi,c}(x) = \min\left\{ |i| : i \in \M, C^{\{i\}, \psi}(x) \leq C^{\Nm, \psi}(x)+c   \right\} \p
\]
When $\M = \Nm$, we will abbreviate with $\s^{\Nm,\psi,c}$. \label{definition:index}
\end{definition}
Koppel and Atlan's treatment \cite{koppelSoph1988,koppel1991almost}, where the name \emph{sophistication} originates, follows this basic logic, although it contains idiosyncracies like the use of monotonic models, and an extension to infinite strings. As the subsequent history of sophistication has discarded these, we will not discuss them here.

In \cite{antunes2009sophistication,antunes2013sophistication} Koppel's principle is limited to finite strings, with the total functions as a model class. The definition used is similar to $s_\text{index}^{\T, \C,\psi,c}$, with one exception: the total complexity of a witness $(i,y)$ is measured as $|i|+|y|$ without the cost of delimiting the two. This difference is not relevant to the point we make in this section. The restriction to total classes is a common approach, perhaps used because it helps to avoid underfitting, which we will discuss in the next section.

\begin{lemma}
Let $\s^\psi$ denote any index sophistication with respect to numbering $\psi$.
There are acceptable numberings $\psi$ and $\xi$ such that for all $x$:\\
\-\hspace{2.8em} $|\s^\psi(x) - \s^\xi(x)|\geq \tfrac{1}{2}\min\{\s^\psi(x),\s^\xi(x)\}$.
\end{lemma}
\begin{proof}
Let $s_i$ be the string consisting of $2^{i}$ zeroes followed by a one. Define $\psi, \xi$ such that $\psi_j(x) = \phi_i(x)$ for $j = s_{2i}$ and $\xi_j(x) = \phi_i(x)$ for $j = s_{2i+1}$, with all other functions returning $\infty$ for all inputs. Choose any $x$ and assume w.l.o.g. that $\s^\psi(x)\le s^\xi(x)$. By construction, we have $2\s^\psi(x)  \leq \s^\xi(x)$.
\end{proof}
Thus, the length of the index is a very poor indicator of model complexity. For a robust measure, define the complexity of functions as in \cite{grunwald2004shannon,vitanyi2004meaningful} by $C^{\M,\psi}(f) = \min\{C^{\M,\psi}(i):\phi_i=f\}$.
Lemma~\ref{lemma:invariance} in the appendix shows that $C^\C(f)$ and $C^\K(f)$ are invariant: they are objective properties of a function.

Note that such perversely inefficient numberings are not an issue for Kolmogorov complexity. If our UTM uses an inefficient numbering, we can switch to a more efficient one at only a constant cost, keeping the complexity the same, up to a constant. For sophistication, however the choice of numbering is crucial.

There are two ways to use the model complexity $C^\M(\phi_i)$ for more robust attempts to define sophistication. Confusingly, both are used in the literature. First, we can measure the complexity of the \emph{model} using $C^{\M,\psi}(\phi_i)$, which is then the size of the first part of a two-part code describing the data. For instance, this approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information} and in most of the present paper.

Second, we can stick to using the length of the index as the measure of sophistication, but restrict the allowed numberings. This approach is taken by Adriaans in \cite{adriaans2012facticity}, which defines \emph{facticity} as $\s_\text{index}^{\C,\psi,0}$, but only allows numberings that are \emph{faithful}: $\forall i \exists j : \psi_i=\psi_j, |j|\le C^{\C}(\psi_j)+c$, for some constant $c$.

We prove in the appendix that---contrary to Adriaans' suggestion---there actually do exist faithful, acceptable numberings (Lemma~\ref{lemma:faithful-numberings}). However even choosing a faithful index is not quite enough. The Kolmogorov complexity uses representations of the form $\bar\imath y$, where the bar denotes some straightforward prefix encoding to delimit the model description $i$ from its input $y$. If we define a second prefix encoding $\tilde{\imath}$, with $|\bar\imath|-|\tilde\imath|$ unbounded, we can define a second representation $\bar u \tilde \imath y$, at a constant overhead $|\br{u}|$, and gain more than $|\br{u}|$ for sufficiently complex strings, resulting in a bounded sophistication.

We continue with a sophistication that avoids these issues. Let $C'^{\M, \psi} = C^\K(\psi_i) + min\{|y| : \psi_i(y) = x\}$. We define sophistication as:

\begin{definition}[Sophistication]
\[
\s^{\M,\Nm,\psi,c}(x)=\min\left\{C^\K(\phi_i):i\in\M,C'^{\{i\},\psi}(x)\le C^{\Nm,\psi}(x)+c\right\}.
\]
\end{definition}

@@

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. We will investigate this balance, starting with the variant $\s^{\K,\psi,c}$. This variant itself is not used anywhere in the literature, but it helps to illustrate the issues we wish to discuss.

$\K$ has representations with all information except a constant in the input and representations with all information in the model. The downside to this balance is that it becomes easy to show a lack of invariance: by tweaking the numbering so that models in a specific subset $\M'\subset\M$ become cheaper to represent by an arbitrary amount relative to others, so that a model in $\M'$ always determines the sophistication.

This means that for model classes with a universal model, we can choose our numbering to make $S^\K$ either bounded, or equal to the complexity:

\begin{theorem}[Underfitting]
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M, x : C'^{\{u\}}(x) \leq C'^{\{i\}}(x) + c$. Then, for some numberings $\s^{\M,\psi,c}$ is bounded. \label{theorem:underfitting}
\end{theorem}

This result is well known and many treatments avoid it by restricting the model class. Less well known is perhaps that the same holds in the other direction: for some numberings the singletons always determine the sophistication.

\begin{theorem}[Overfitting]
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\M(x)-S^{\M,\psi,c}(x)\leq c$.\label{theorem:overfitting}
\end{theorem}
The proofs of both theorems rely on a simple principle: for certain $\M' \subset \M$, there exist numberings which have the effect of penalizing $C^\K(\phi_i)$ for any model outside $\M'$ by an arbitrary constant amount. We can use this to effectively `push' these models outside of the range of candidates, ensuring that, under this numbering, a model in $\M'$ always determines the sophistication.

For these theorems, this means setting $\M' = \{u\}$, with $\phi_u$ a universal model, or letting $\M'$ be the singletons. The requirements for $\M'$ are strict, and slightly complex. The following lemma gives a set of sufficient conditions.

\begin{restatable}{lemma}{coolone}
\label{lemma:thecoolone}
  Let $\M$ be any model class, let $\X$ be any set of binary strings and let $D:\B\to\N$ be a computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\phi$. Let $\M'= \tn{range}(D)$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(1) $\forall_{m\in\M'}:\min\{|p|:\phi_{D(p)}=\phi_m\}\le C^{\K,\phi}(\phi_m)+c$\\
\-\hspace{1cm}(2) $\forall_{x\in\X}:C'^{\M',\phi}(x)-C'^{\M,\phi}(x)\le c$.\\
Then there is an enumeration $\psi$ of the partial recursive functions such that $S^{\M,\psi,c}(x) = |\br{0}|+S^{\M',\phi,c}(x)$ for all $x\in\X$.
\end{restatable}

\begin{proof}
We define the numbering $\psi$ as follows:
\[
\psi_0(p) = 1^r 0 D(p), \;\;\;
\psi_{1^r0i}(p) = \phi_i(p), \;\;\;
\psi_j(\cdot) = \infty \;\text{if $j$ contains no zeroes.}
\]
We will show that under the $\psi$, the best representation for $x$ with $m\in\M'$ is always better than the best using some $m\notin\M'$. First suppose $m\in\M'$. Then
\begin{align*}
   C^{\K,\psi}(\phi_m) &=\min\{|\bar\jmath q|:\psi_{\psi_j(q)}=\phi_m\} \leq\min\{|\bar0 q|:\psi_{\psi_0(q)}=\phi_mm\} \\ 
   &=\min\{|\bar0 q|:\psi_{1^r0 D(q)}=\phi_m\} =\min\{|q|:\phi_{D(q)}=\phi_m\} + |\bar 0|\\
   &\leq C^{\K,\phi}(\phi_m)+c+|\br{0}|,
\end{align*}
where the last inequality uses assumption (1).
